Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-11 inc/photocathode/setup/setup.tex
citations are now preceeded by non-breaking space (~)

%Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED) \cite{sciaini_femtosecond_2011}, ultrafast transmission electron microscopy (UEM)\cite{king_ultrafast_2005}, and pulsed X-ray free electron lasers \cite{dowell_cathode_2010}.
The spatial quality of the sources for UEM and UED are often characterized in terms of their emitted electron beam's normalized transverse rms emittance \cite{dowell_quantum_2009,jensen_emittance_2010}; $\varepsilon_{\smallT} = \Delta p_{\smallT} \Delta x / (m c)$, where $\Delta p_{\smallT} = \sqrt{ \langle p_{\smallT}^2 \rangle }$ is the rms transverse momentum of the beam, $\Delta x = \sqrt{ \langle x^2 \rangle }$ is its size (cylindrical symmetry assumed), and $m$ is the electron rest mass and $c$ is the speed of light in vacuum.
This quantity is conserved under free-space propagation and perfect lensing, but may increase due to lens aberrations \cite{oshea_reversible_1998}.
As I have discussed in Chapter \ref{chap:considerations}, an electron beam (or in the case of UEM, a pulse) with a lower transverse emittance will result in higher quality images, however since this quantity is conserved, the only chance to minimize this quantity occurs upon emission.
As a reduction in the source size $\Delta x_0$ is limited either by the attainable diffraction-limited incident laser pulse focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field) \cite{valfells_effects_2002}, significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness \cite{berger_dc_2009}) may only be possible by decreasing the initial rms transverse momentum $\Delta p_{\smallT}$ of the electron source --- an intrinsic property of the emission source \cite{dowell_quantum_2009,jensen_emittance_2010}.

@@

\ref{fig:transverse-measurement} depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_{\smallT}$) of the laser-driven electron sources.
Though many of the experimental components have been described in detail in Chapter \ref{chap:prototype}, for convenience I will briefly summarize them again here.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator \cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface \cite{berger_dc_2009}.
The electron gun design is based on the that of Togawa \textit{et al.} \cite{berger_dc_2009,togawa_ceb6_2007} and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.

>>>>

%Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED)~\cite{sciaini_femtosecond_2011}, ultrafast transmission electron microscopy (UEM)~\cite{king_ultrafast_2005}, and pulsed X-ray free electron lasers~\cite{dowell_cathode_2010}.
The spatial quality of the sources for UEM and UED are often characterized in terms of their emitted electron beam's normalized transverse rms emittance~\cite{dowell_quantum_2009,jensen_emittance_2010}; $\varepsilon_{\smallT} = \Delta p_{\smallT} \Delta x / (m c)$, where $\Delta p_{\smallT} = \sqrt{ \langle p_{\smallT}^2 \rangle }$ is the rms transverse momentum of the beam, $\Delta x = \sqrt{ \langle x^2 \rangle }$ is its size (cylindrical symmetry assumed), and $m$ is the electron rest mass and $c$ is the speed of light in vacuum.
This quantity is conserved under free-space propagation and perfect lensing, but may increase due to lens aberrations~\cite{oshea_reversible_1998}.
As I have discussed in Chapter \ref{chap:considerations}, an electron beam (or in the case of UEM, a pulse) with a lower transverse emittance will result in higher quality images, however since this quantity is conserved, the only chance to minimize this quantity occurs upon emission.
As a reduction in the source size $\Delta x_0$ is limited either by the attainable diffraction-limited incident laser pulse focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field)~\cite{valfells_effects_2002}, significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness~\cite{berger_dc_2009}) may only be possible by decreasing the initial rms transverse momentum $\Delta p_{\smallT}$ of the electron source --- an intrinsic property of the emission source~\cite{dowell_quantum_2009,jensen_emittance_2010}.

@@

\ref{fig:transverse-measurement} depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_{\smallT}$) of the laser-driven electron sources.
Though many of the experimental components have been described in detail in Chapter \ref{chap:prototype}, for convenience I will briefly summarize them again here.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator~\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface~\cite{berger_dc_2009}.
The electron gun design is based on the that of Togawa \textit{et al.}~\cite{berger_dc_2009,togawa_ceb6_2007} and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.

### 2013-04-11 inc/photocathode/single_photon/single_photon.tex
citations are now preceeded by non-breaking space (~)

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect'' \cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it becomes free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$; for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as, %TODO ref
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material paramenter (typically of the order of $\sim 10^{-4} \text{eV}^{-2}$ %TODO ref
), and $R$ is the photocathode surface reflectivity, which is itself a function of the polarization, frequency and angle of the incident light.
%TODO calculation of $\eta_{PE}$?

@@

An example of single-photon photoemission is presented in \ref{fig:single_photon_tungsten}, wherein a sample of polished tungsten (W), having a work function $\Phi \approx $ 4.3-4.5 eV \cite{yen_thermally_1980}, is used as a photocathode.
When paired with our $\hbar \omega = $ 4.75 eV photon energy laser (see Section \ref{sec:laser}) the excess photoemission energy $\Delta E \approx $ 0.35 eV.
The linear dependence of the electron yield (circles) on the laser pulse energy is indicative of a single photon process, as expected since each absorbed photon may eject one electron.
In this experiment, the yield is determined by integrating the total counts recorded on the CCD detector and therefore is only a relative measurement.
The fact that the Fourier plane spot-size (squares), directly related the initial transverse momentum variance $\Delta p_{\smallT}$, is unchanged over the full range of laser power is also expected, since by $\Delta p_{\smallT} = \sqrt{ m \Delta E / 3 } $ \cite{dowell_quantum_2009}, $\Delta p_{\smallT}$ only depends on $\Delta E$.

In order to improve performance, one must be able to minimize $\Delta p_{\smallT}$ without negatively affecting the photoemission efficiency, and thus electron yield.
For single-photon photoemission, although the polarization-dependent reflectivity can be tweaked to improve performance slightly \cite{berger_dc_2009}, due to the quadratic dependence of the efficiency on excess photoemission energy \cite{shalaev_electron_1994}, any reduction in $\Delta p_{\smallT}$ gained by reducing $\Delta E$ will be mitigated by a large drop in electron yield.
For this reason, as it stands currently and as we have already seen in Section \ref{sec:must_reduce_transverse_momentum}, flat metal photocathodes employing single-photon photoemission are not likely to be useful for UEM.
However, as I will show in Section \ref{sec:photocathode_future}, this does not quite tell the complete story.

>>>>

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect''~\cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it becomes free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$; for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as, %TODO ref
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material paramenter (typically of the order of $\sim 10^{-4} \text{eV}^{-2}$ %TODO ref
), and $R$ is the photocathode surface reflectivity, which is itself a function of the polarization, frequency and angle of the incident light.
%TODO calculation of $\eta_{PE}$?

@@

An example of single-photon photoemission is presented in \ref{fig:single_photon_tungsten}, wherein a sample of polished tungsten (W), having a work function $\Phi \approx $ 4.3-4.5 eV~\cite{yen_thermally_1980}, is used as a photocathode.
When paired with our $\hbar \omega = $ 4.75 eV photon energy laser (see Section \ref{sec:laser}) the excess photoemission energy $\Delta E \approx $ 0.35 eV.
The linear dependence of the electron yield (circles) on the laser pulse energy is indicative of a single photon process, as expected since each absorbed photon may eject one electron.
In this experiment, the yield is determined by integrating the total counts recorded on the CCD detector and therefore is only a relative measurement.
The fact that the Fourier plane spot-size (squares), directly related the initial transverse momentum variance $\Delta p_{\smallT}$, is unchanged over the full range of laser power is also expected, since by $\Delta p_{\smallT} = \sqrt{ m \Delta E / 3 } $~\cite{dowell_quantum_2009}, $\Delta p_{\smallT}$ only depends on $\Delta E$.

In order to improve performance, one must be able to minimize $\Delta p_{\smallT}$ without negatively affecting the photoemission efficiency, and thus electron yield.
For single-photon photoemission, although the polarization-dependent reflectivity can be tweaked to improve performance slightly~\cite{berger_dc_2009}, due to the quadratic dependence of the efficiency on excess photoemission energy~\cite{shalaev_electron_1994}, any reduction in $\Delta p_{\smallT}$ gained by reducing $\Delta E$ will be mitigated by a large drop in electron yield.
For this reason, as it stands currently and as we have already seen in Section \ref{sec:must_reduce_transverse_momentum}, flat metal photocathodes employing single-photon photoemission are not likely to be useful for UEM.
However, as I will show in Section \ref{sec:photocathode_future}, this does not quite tell the complete story.

### 2013-04-11 inc/photocathode/two_photon_thermionic/two_photon_thermionic.tex
citations are now preceeded by non-breaking space (~)

Two-photon assisted thermionic emission (2PTE) occurs when two photons provide nearly enough energy to emit a cold electron, and therefore a hotter electron, in the tail of the Fermi distribution, may have enough energy to be emitted \cite{yen_thermally_1980}.
Of course this is even less efficient than direct two-photon photoemission, but as three-photon photoemission goes as $I^3$, 2PTE will still be the dominant photoemission mechanism.
For thermionic emission from electrons at a temperature $T_e$, the rms transverse momentum $\Delta p_{\smallT}$ is given by 
\begin{equation}
  \Delta p_{\smallT} = \sqrt{m k_B T_e} \,\text{,}
\end{equation}
where $k_B$ is the Boltzmann constant \cite{dowell_quantum_2009,jensen_emittance_2010}.
Since a high laser intensity is required to drive 2PTE, the laser itself will heat the electrons, therefore, if the laser pulse duration is less than the thermalization time of the electron distribution, %TODO check this!
we expect the measured Fourier plane spot size to grow as incident laser pulse energy increases.

@@

The quadratic dependence of the electron yield on pulse energy clearly indicates a two-photon emission process.
This is expected since an excitation energy of two green photons ($\hbar \omega$ = 2.37eV) will be required to overcome the reported $\Phi$ = 4.69eV effective work function of a thick gold film contaminated with adsorbed water \cite{monjushiro_ultraviolet_1991} or ensure that at least the tail of the electron Fermi distribution has sufficient energy to overcome the 5.1($\pm$0.1)eV work function of a clean Au surface \cite{eastman_photoelectric_1970}.
Moreover, when $2\hbar \omega \approx \Phi$, the dependence of the yield on the square of the electron's excess energy above the work function \cite{monjushiro_ultraviolet_1991} ensures that predominantly only the high energy (Boltzmann) tail of the Fermi distribution contributes to the emission.
Consequently, the observed increase in $\Delta p_{\smallT}$ with incident pulse energy must be due to a heating effect; specifically, laser heating of the free electron Fermi gas as two-photon excitation is instantaneous and the $\sim$200fs green laser pulse duration excludes any coupling to the lattice which occurs on the time scale of a few picoseconds \cite{chen_semiclassical_2006}.
The solid red line in \ref{fig:gold-emission} is the predicted variation derived using a simple zero free parameter model of this effect.
Using the known optical properties of gold \cite{johnson_optical_1972}, which give a reflectance $R_p$ = 0.53 and an absorption depth of 20nm, and the temperature-dependent heat capacity of a free electron Fermi gas, the model evaluates the average temperature of the laser-heated electron gas that may be two-photon excited assuming that only electrons within a few nanometers of the surface can be emitted, as the mean free path for electrons in Au is $\sim$4nm \cite{seah_quantitative_1979}.
This average electron temperature $T_e$ is then used in the AG model simulation of the experiment (\ref{fig:transverse-measurement}) to determine the expected Fourier plane spot size on the YAG scintillator; that is, using an initial $\Delta p_{\smallT} = \sqrt{m k_B T_e}$.  The dashed red lines indicate the expected trends for a $\pm$10\% error in the laser spot size $\Delta x_0$ incident on the gold cathode surface.
The close agreement between the data and the simulation strongly supports our interpretation of the emission mechanism.

>>>>

Two-photon assisted thermionic emission (2PTE) occurs when two photons provide nearly enough energy to emit a cold electron, and therefore a hotter electron, in the tail of the Fermi distribution, may have enough energy to be emitted~\cite{yen_thermally_1980}.
Of course this is even less efficient than direct two-photon photoemission, but as three-photon photoemission goes as $I^3$, 2PTE will still be the dominant photoemission mechanism.
For thermionic emission from electrons at a temperature $T_e$, the rms transverse momentum $\Delta p_{\smallT}$ is given by 
\begin{equation}
  \Delta p_{\smallT} = \sqrt{m k_B T_e} \,\text{,}
\end{equation}
where $k_B$ is the Boltzmann constant~\cite{dowell_quantum_2009,jensen_emittance_2010}.
Since a high laser intensity is required to drive 2PTE, the laser itself will heat the electrons, therefore, if the laser pulse duration is less than the thermalization time of the electron distribution, %TODO check this!
we expect the measured Fourier plane spot size to grow as incident laser pulse energy increases.

@@

The quadratic dependence of the electron yield on pulse energy clearly indicates a two-photon emission process.
This is expected since an excitation energy of two green photons ($\hbar \omega$ = 2.37eV) will be required to overcome the reported $\Phi$ = 4.69eV effective work function of a thick gold film contaminated with adsorbed water~\cite{monjushiro_ultraviolet_1991} or ensure that at least the tail of the electron Fermi distribution has sufficient energy to overcome the 5.1($\pm$0.1)eV work function of a clean Au surface~\cite{eastman_photoelectric_1970}.
Moreover, when $2\hbar \omega \approx \Phi$, the dependence of the yield on the square of the electron's excess energy above the work function~\cite{monjushiro_ultraviolet_1991} ensures that predominantly only the high energy (Boltzmann) tail of the Fermi distribution contributes to the emission.
Consequently, the observed increase in $\Delta p_{\smallT}$ with incident pulse energy must be due to a heating effect; specifically, laser heating of the free electron Fermi gas as two-photon excitation is instantaneous and the $\sim$200fs green laser pulse duration excludes any coupling to the lattice which occurs on the time scale of a few picoseconds~\cite{chen_semiclassical_2006}.
The solid red line in \ref{fig:gold-emission} is the predicted variation derived using a simple zero free parameter model of this effect.
Using the known optical properties of gold~\cite{johnson_optical_1972}, which give a reflectance $R_p$ = 0.53 and an absorption depth of 20nm, and the temperature-dependent heat capacity of a free electron Fermi gas, the model evaluates the average temperature of the laser-heated electron gas that may be two-photon excited assuming that only electrons within a few nanometers of the surface can be emitted, as the mean free path for electrons in Au is $\sim$4nm~\cite{seah_quantitative_1979}.
This average electron temperature $T_e$ is then used in the AG model simulation of the experiment (\ref{fig:transverse-measurement}) to determine the expected Fourier plane spot size on the YAG scintillator; that is, using an initial $\Delta p_{\smallT} = \sqrt{m k_B T_e}$.  The dashed red lines indicate the expected trends for a $\pm$10\% error in the laser spot size $\Delta x_0$ incident on the gold cathode surface.
The close agreement between the data and the simulation strongly supports our interpretation of the emission mechanism.
